Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-08-20 writeups/Su_IGW_break.tex
updated writeup for Daniel revision 190820

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We furthermore neglect temperature variations and
assume a barotropic equation of state $P(\rho, T) = P(\rho)$ as a first
approximation. As we are interested in dynamics far from the center of the WD,
we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho}(x, z) = \overline{\rho}_0 e^{-z/H}$
for some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes).

@@

In practice, it is convenient to introduce the coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ \citep[e.g.][]{lecoanet_anel}. This both
identically enforces $\rho > 0$ and eliminates the stiff term
$\frac{\bm{\nabla}P}{\rho}$. We also define reduced pressure $\varpi =
\frac{P}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}'}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\uv{z} &= 0.
    \end{align}
\end{subequations}
In the new coordinates, hydrostatic equilibrium corresponds to $\Upsilon = 0,
\overline{\varpi} = gH$.

@@

\subsection{Analytical Properties: Linear}

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u}' \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime is given by \citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}\cos\p{\p{k_{x}x + k_{z}z - \omega t}},
        \label{eq:k0z_sign}
\end{equation}
where $A$ is the amplitude and $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed $c_s \to \infty$, in
which the \emph{Brunt-V\"ais\"al\"a frequency}
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$ \citep[see
e.g.][]{sutherland0}.

@@

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{p} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        \citep[e.g.][]{drazin,sutherland1}. Note that for an upwards
        propagating IGW $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\uv{z}$
        direction can be computed. Since the linear solution consists of terms
        separable as $f(x, z, t) = f(z)\cos\p{k_{x}x - \omega t}$, $x$
        averaging and time averaging are equivalent. $x$ averaging is preferred
        both since periodic boundary conditions pick a natural averaging length
        and since it does not introduce extra temporal smoothing. We may then
        write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging. Note that indeed for an
        upwards propagating IGW $k_z < 0, k_x > 0$, we have $F > 0$.
\end{itemize}

\subsection{Wave Generation: Analytical Analysis}

Since we are only concerned with the behavior of the wave at high $z$ where wave
breaking occurs, we generate waves at low $z$ in the most numerically convenient
way. As long as physically represesentative, upwards-propgaating IGW are
generated and wave dynamics are only examined far from the forcing region, the
exact excitation mechanism is not important for our conclusions.

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain\footnote{Interfacial forcing is numerically
unstable when using a Chebyshev polynomial basis in the $z$ direction and
impossible when using a Fourier basis.}. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are damped away by the damping layers described in
\autoref{ss:damping}.

@@

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two plane wave solutions, above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z$), and we obtain
\begin{align}
    u_{z}'&(x, z, t) ={} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upwards-propagating IGW wavetrain excited deep
in the atmosphere.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\bm{u}' \cdot \bm{\nabla}$. However, since $\abs{\bm{u}'} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement of the perturbation $\bm{\xi}' =
\frac{\bm{u}'}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \abs{\bm{\xi}'} \abs{\bm{k}} \sim \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for where $\overline{U} = \omega / k_x$ the critical layer
condition is fulfilled (\autoref{eq:u_c} and discussion in
\autoref{ss:crit_layer}), we obtain $\abs{\frac{u_{x}'k_{x}}{\omega}} =
\abs{\xi_x k_{x}} = \abs{\xi_z k_z} \sim 1$, equal to the nonlinearity criterion
\autoref{eq:nl}. This condition predicts where the wave's self-acceleration
alone produces a critical layer.

\subsection{Formation and Dynamics of Critical Layer}\label{ss:crit_layer}

[TODO rewrite more clearly, it's just copy pasted together right now]

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries \citep{zahn75,gn89} as well as the
quasi-biennial oscillation \citep{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions \citep{drazin}, causing energy transfer out of the IGW
to daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow \citep{klostermeyer}.

In accordance with literature and the simulation setup, we will assume momentum
redistribution is inefficient (see \autoref{ss:disp} for analysis of this
assumption). The mean flow will then grow in place until it reaches the
horizontal phase velocity of the parent IGW\@. Then, a critical layer forms at
which the frequency of the incident IGW is Doppler-shifted to zero. The
interaction of the parent IGW with the mean flow was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero viscosity
\citep{hazel}, but weakly nonlinear theory \citep{brown_stewartson} and fully
nonlinear simulations \citep{winters1994} suggest that nonlinear effects can
induce reflection.

A purely horizontal shear flow $\overline{U}(z) \uv{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. If anywhere $\overline{U}(z) = \overline{U}_c$ where
\begin{equation}
    \overline{U}_c \equiv \omega / k_x,\label{eq:u_c}
\end{equation}
then the horizontal wavenumber of the linear wave in the fluid frame vanishes
and critical behavior is observed. In a linear theory, both inviscid
\citep{booker_bretherton} and viscous \citep{hazel}, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R}_A &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T}_A &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the mean flow cannot exceed $\overline{U}_c$ the
horizontal phase velocity of the incident wave, the critical layer must instead
propagate downwards to accommodate the incident momentum. The total horizontal
momentum of the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta F = 0.
\end{equation}
$\Delta F$ is the total injected horizontal momentum to the flow. We next make a
simple two-zone approximation $\overline{U}(z > z_c) = \overline{U}_c,
\overline{U}(z < z_c) = 0$, assuming all horizontal momentum flux goes into
critical layer propagation. Then, the conservation equation becomes simple
differential equation
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If the entirety of the horizontal momentum flux carried by the IGW wavetrain is
absorbed by the critical layer, then $\Delta F = -F'$ is constant in time.
For constant $\Delta F$ in time and $\rho \approx \overline{\rho}$, $z_c(t)$ has
analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta F}
        .\label{eq:zc_sol}
\end{equation}

@@

In a realistic WD, IGW are excited very deep in the WD and propagate a
substantial number of density scale heights before reaching the breaking zone.
This propagation is very nearly exactly linear until just a few scale heights
below the breaking region. In the interest of computational feasibility, we
restrct the simulation domain to the upper WD fluid, of much smaller vertical
extent than the full physical domain. To further decrease computational cost, we
restrict ourselves to 2D, plane-parallel IGW rather than full 3D stellar
g-modes. While it is well-known wave breaking is a 3D process
\citep{klostermeyer,winters1994}, the dynamical effect of the breaking process
is likely to be similar in 2D.

@@

The geometry of our simulation domain is fixed by one further parameter: $z_0$
the forcing location. We use $z_0 = 0.2L_z$ to force sufficiently far from the
lower damping zone and permit sufficient room for the upwards-moving wave to
grow at $\propto e^{z/2H}$. We choose $L_z = 10H$ to give $\sim e^3$ apmlitude
growth between the damping zones. Finally, we want similar grid spacing
$\frac{L_x}{N_x} \sim \frac{L_z}{N_z}$, guided by the intuition that turbulence
is approximately isotropic. For computational savings, we fix $L_x = 4H < L_z$,
a compromise between using fewer $N_x$ modes and using as large of an $L_x$ as
possible (in stars, $L_x \sim 2\pi R$ the circumference). We then fix ratio $N_z
/ N_x = 4$.

@@

The nonlinear terms in the above equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
dissipation, artificial dissipation must be added. To ensure the dissipitive
system conserves horizontal momentum exactly, we begin by adding dissipitive
terms to the flux-conservative form of the Euler fluid equations
\autoref{se:nl_orig} (we use stress tensor $\tau_{ij} = P\delta_{ij}$):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}' &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u}' - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}') + \bm{\nabla} \cdot (\rho \bm{u}' \bm{u}'
            + \mathrm{diag}(\rho \varpi)
            - \nu \rho \bm{\nabla}\bm{u})
            + \rho g \uv{z} &= 0.
    \end{align}
\end{subequations}
The same $\nu$ is used for both the diffusive and viscous term, though this is
not required. Since the dissipation is not physical and is purely used for
numerical stability, we choose it such that hydrostatic equilibrium is not
modified (hence $\nu$ acts only on $\rho - \overline{\rho}$).
% Some algebraic manipulation to re-cast it in the form of \autoref{se:nl_var}
% gives
% \begin{subequations}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u} &= 0,\\
%         \partial_t \Upsilon + \p{\bm{u} \cdot \bm{\nabla}} \Upsilon -
%             \frac{u_z}{H} - \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
%             \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
%             \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}} &= 0,\\
%         \partial_t \bm{u} + \p{\bm{u} \cdot \bm{\nabla}}\bm{u} +
%             \bm{\nabla} \varpi + \varpi \bm{\nabla} \Upsilon - \nu \nabla^2
%             \bm{u} + \bm{u} \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
%             \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
%             \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}}&{}\nonumber\\
%         - 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}\bm{u} -
%             \frac{1}{H}\partial_z \bm{u}} - \frac{\varpi'}{H} &= 0.
%     \end{align}
% \end{subequations}
One last consideration we found necessary was masking out nonlinear terms in the
forcing zone with a similar form to \autoref{eq:Gamma}. In the absence of this
mask, a strong mean flow localized to the forcing zone developed. The mask used
was
\begin{equation}
    \Gamma_{NL}(z) = \frac{1}{2}\s{2
        + \tanh \frac{z - (z_0 + 8\sigma)}{\sigma}
        - \tanh \frac{z - z_B}{\sigma}}.
\end{equation}

>>>>

We consider a incompressible, isothermal fluid, representative of degenerate
matter in WDs. We use barotropic equation of state $P(\rho, T) = P(\rho)$ as a
first approximation. As we are interested in dynamics far from the center of the
WD, we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho} = \overline{\rho}_0 e^{-z/H}$ for
some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes). Finally, we
consider a 2D fluid for computational feasibility: while it is well-known wave
breaking is a 3D process \citep{klostermeyer,winters1994}, the dynamical effect
of the breaking process is likely to be similar in 2D \citep{barker_ogilvie}.

@@

In practice, it is convenient to introduce the coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ \citep[e.g.][]{lecoanet_anel}. This both
identically enforces $\rho > 0$ and eliminates the stiff term
$\frac{\bm{\nabla}P}{\rho}$. We also define reduced pressure $\varpi =
\frac{P}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0.
    \end{align}
\end{subequations}
In the new coordinates, hydrostatic equilibrium corresponds to $\Upsilon = 0,
\overline{\varpi} = gH$.

@@

\subsection{Linear Analysis}

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u}' \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime is given by \citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:k0z_sign}
\end{equation}
where $A$ is the amplitude and $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed $c_s \to \infty$, in
which the \emph{Brunt-V\"ais\"al\"a frequency}
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$ \citep[see
e.g.][]{sutherland0}.

@@

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\p{ \p{k_{z}H}^{-2} }
        \approx 0$. In the Boussinesq approximation where density stratification
        outside of $N^2$ is completely neglected, the phase and group velocity
        are exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

\subsection{Wave Generation}

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are damped away by the damping layers described in
\autoref{ss:damping}.

@@

In the linearized system, the effect of this forcing can be solved analytically
to be
\begin{align}
    u_{z}'&(x, z, t) ={} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\cos\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere.

@@

Above, we neglected advective terms $\pd{}{t} \gg \bm{u}' \cdot \bm{\nabla}$.
However, since $\abs{\bm{u}'} \propto e^{z/2H}$ in an infinite domain all IGWs
eventually grow to nonlinear amplitudes and break. Nonlinear wave breaking is
expected to be important in WDs \citep{fullerI,fullerII}. We can
estimate the height of wave breaking using $\frac{\abs{\bm{u}'}
\abs{\bm{k}}}{\omega} \sim 1$. This can be rewritten using the Lagrangian
displacement $-i\omega \bm{\xi}' = \bm{u}'$:
\begin{equation}
    \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

An equivalent estimate given by \citep{eliassen_palm_cite,sutherland0} is that
IGWs break when the wave-induced mean flow is equal to the horizontal phase
velocity
\begin{equation}
    \overline{U}(z) = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}} \sim
        \frac{\omega}{k_x},\label{eq:mean_flow}
\end{equation}
where $\bar{U}(z) \equiv \ev{u_x}_x$ is the mean flow of the fluid.
\autoref{eq:mean_flow} provides a leading-order estimate for $\bar{U}(z)$ which
may further evolve over time (see \autoref{ss:crit_layer}); for a general mean
flow, \autoref{eq:mean_flow} also defines the criterion for a \emph{critical
layer}. This wave-induced mean flow is analogous to Stokes' drift for surface
waves and can be derived by considering the propagation of $F$
(\autoref{eq:F_def}) into a medium at rest. Evaluating \autoref{eq:mean_flow} yields
agreement with \autoref{eq:nl}.

\subsection{Critical Layer Dynamics}\label{ss:crit_layer}

% critical layer does form
% previous studies of fluid interaction @ critical layer

The detailed onset of wave breaking has been laid out in a few key papers
\citep{drazin, klostermeyer, winters1994}, where instabilities transfer momentum
out of the IGW through daughter modes into the mean flow of the fluid. When
viscosity is sufficiently weak (see \autoref{ss:disp} for a discussion of this
assumption in WDs), a non-uniform shear flow and eventually a critical layer
develop. The further evolution of the critical layer with the incident IGW is
thought to be responsible for tidal synchronization in stellar binaries
\citep{zahn75,gn89} as well as the quasi-biennial oscillation
\citep{lindzen_qbo}.

The behavior of an incident IGW upon a critical layer was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption of the IGW\@. In particular, the incident wave has amplitude
reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R}_A &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T}_A &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the height of the critical
layer $z_c$. In the $\mathrm{Ri} \gg 1$ limit, $\mathcal{R}, \mathcal{T} \ll 1$
and the incident wave is completely absorbed to good approximation.

This result extends to viscous fluids \citep{hazel}, but weakly nonlinear theory
\citep{brown_stewartson} and numerical simulations \citep{winters1994} suggest
that nonlinear effects significantly enhance reflection and transmission. These
studies are not directly applicable to our problem where a continuous train of
IGW are excited.

Instead, we mayanalyze horizontal momentum transfer at the critical layer. Any
incident horizontal momentum flux absorbed by the fluid must manifest as
additional horizontal momentum of the shear flow. Since the mean flow cannot
exceed $\overline{U}_c \equiv \omega/k_x$ the horizontal phase velocity of the
incident wave, the critical layer must instead propagate downwards to
accommodate the incident momentum flux. The total horizontal momentum of the
shear flow obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta F = 0,
\end{equation}
where $\Delta F$ is the absorbed horizontal momentum. Assuming $\overline{U}(z >
z_c) = \overline{U}_c, \overline{U}(z < z_c) = 0$, the conservation equation
becomes differential equation for $z_c(t)$
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If the entirety of the horizontal momentum flux carried by the IGW wavetrain is
absorbed by the critical layer, then $\Delta F = -F'$ is constant in time.
Finally, for constant $\Delta F$ in time and $\rho \approx \overline{\rho}$,
$z_c(t)$ has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta F}
        .\label{eq:zc_sol}
\end{equation}

@@

In a realistic WD, IGW are excited very deep in the WD and propagate a
substantial number of density scale heights before reaching the breaking zone.
This propagation is very nearly exactly linear until just a few scale heights
below the breaking region. In the interest of computational feasibility, we
restrct the simulation domain to the upper WD fluid, of much smaller vertical
extent than the full physical domain. To further decrease computational cost, we
restrict ourselves to 2D, plane-parallel IGW rather than full 3D stellar
g-modes.

@@

The geometry of our simulation domain is fixed by one further parameter: $z_0$
the forcing location. We use $z_0 = 0.2L_z$ to force sufficiently far from the
lower damping zone and permit sufficient room for the upward moving wave to
grow at $\propto e^{z/2H}$. We choose $L_z = 10H$ to give $\sim e^3$ apmlitude
growth between the damping zones. Finally, we want similar grid spacing
$\frac{L_x}{N_x} \sim \frac{L_z}{N_z}$, guided by the intuition that turbulence
is approximately isotropic. For computational savings, we fix $L_x = 4H < L_z$,
a compromise between using fewer $N_x$ modes and using as large of an $L_x$ as
possible (in stars, $L_x \sim 2\pi R$ the circumference). We then fix ratio $N_z
/ N_x = 4$.

@@

The nonlinear terms in the above equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
dissipation, artificial dissipation must be added. To ensure the dissipitive
system conserves horizontal momentum exactly, we begin by adding dissipitive
terms to the flux-conservative form of the Euler fluid equations
\autoref{se:nl_orig} (we use stress tensor $\tau_{ij} = P\delta_{ij}$):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}' &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u}' - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}') + \bm{\nabla} \cdot (\rho \bm{u}' \bm{u}'
            + \mathrm{diag}(\rho \varpi)
            - \nu \rho \bm{\nabla}\bm{u})
            + \rho g \uv{z} &= 0.
    \end{align}
\end{subequations}
The same $\nu$ is used for both the diffusive and viscous term, though this is
not required. Since the dissipation is not physical and is purely used for
numerical stability, we choose it such that hydrostatic equilibrium is not
modified (hence $\nu$ acts only on $\rho - \overline{\rho}$).

One last consideration we found necessary was masking out nonlinear terms in the
forcing zone with a similar form to \autoref{eq:Gamma}. In the absence of this
mask, a strong mean flow localized to the forcing zone developed. The mask used
was
\begin{equation}
    \Gamma_{NL}(z) = \frac{1}{2}\s{2
        + \tanh \frac{z - (z_0 + 8\sigma)}{\sigma}
        - \tanh \frac{z - z_B}{\sigma}}.
\end{equation}
